---
title: "EDA: Street Trees"
format: html
editor: visual
---

Read in only the columns we need:

```{r}
library(tidyverse)
library(fst)
library(naniar)

# Read only the columns we use
street_trees <- read_csv2("../data/raw/street-trees.csv")
```
In the latest update (as of 2025 March 25) some are missing neighbourhoods!

```{r}
# Count and percentage of missing NEIGHBOURHOOD_NAME
street_trees |> 
  summarise(
    missing_neighbourhood = sum(is.na(NEIGHBOURHOOD_NAME) | NEIGHBOURHOOD_NAME == ""),
    total = n(),
    percent_missing = round(100 * missing_neighbourhood / total, 2)
  )

# Show a few examples
street_trees |> 
  filter(is.na(NEIGHBOURHOOD_NAME) | NEIGHBOURHOOD_NAME == "")
  head(50)
```
It really seems like all of these trees with missing have = `PARK`. Let's confirm.

```{r}
# TRUE if all missing NEIGHBOURHOOD_NAME values have STREET_SIDE_NAME == "PARK"
all_missing_neigh_are_parks <- street_trees |>
  filter(is.na(NEIGHBOURHOOD_NAME) | NEIGHBOURHOOD_NAME == "") |>
  summarise(all_parks = all(STREET_SIDE_NAME == "PARK")) |>
  pull(all_parks)

all_missing_neigh_are_parks
```

Nope. Let's also check whether all the park trees have missing neighbourhoods:

```{r}
# TRUE if all park trees are missing neighbourhood names
all_parks_missing_neigh <- street_trees |>
  filter(STREET_SIDE_NAME == "PARK") |>
  summarise(all_missing = all(is.na(NEIGHBOURHOOD_NAME) | NEIGHBOURHOOD_NAME == "")) |>
  pull(all_missing)

all_parks_missing_neigh
```

Then, let's fill the missing neighbourhood trees to explicitly be "[Missing Neighbourhood]":

```{r}
street_trees <- street_trees |>
  mutate(
    NEIGHBOURHOOD_NAME = ifelse(
      is.na(NEIGHBOURHOOD_NAME) | trimws(NEIGHBOURHOOD_NAME) == "",
      "[Missing Neighbourhood]",
      NEIGHBOURHOOD_NAME
    )
  )
```

Let's make a new column for park trees because those may be interesting:

```{r}
street_trees <- street_trees |>
  mutate(
    PARK_TREE = STREET_SIDE_NAME == "PARK"
  )
```

Filter down to only the columns we need from now on:

```{r}
street_trees <- street_trees |> select(TREE_ID, CIVIC_NUMBER, STD_STREET, GENUS_NAME, SPECIES_NAME,
    COMMON_NAME, NEIGHBOURHOOD_NAME, HEIGHT_RANGE, geo_point_2d, PARK_TREE)
```

Let's examine:

```{r}
head(street_trees, 50)
street_trees |> filter(PARK_TREE == TRUE) |>  head(50)
```

Any lat / lon missingness?

```{r}
# Parse geo_point_2d into lat/lon
street_trees <- street_trees |>
  separate(geo_point_2d, into = c("LATITUDE", "LONGITUDE"), sep = ",\\s*", convert = TRUE)

# Check for missing lat/lon BEFORE filtering
street_trees |> 
  select(LATITUDE, LONGITUDE) |> 
  summarise(
    missing_lat = sum(is.na(LATITUDE)),
    missing_lon = sum(is.na(LONGITUDE)),
    total_missing_either = sum(is.na(LATITUDE) | is.na(LONGITUDE)),
    total_rows = n()
  )

# Visualize missingness
street_trees |> 
  select(LATITUDE, LONGITUDE) |>
  gg_miss_var(show_pct = TRUE)

# View sample rows missing either lat or lon
street_trees_missing_coords <- street_trees |> 
  filter(is.na(LATITUDE) | is.na(LONGITUDE))

street_trees_missing_coords |> 
  head(50)
```
Looks like in the newest update to their dataset, they've fixed the missing lat/lon values. So we don't need this, at least in this release:

```{r}
# Drop rows with missing lat/lon
# street_trees <- street_trees |> 
#   filter(!is.na(LATITUDE) & !is.na(LONGITUDE))
```

Let's do our usual preprocessing steps to clean up the data. We will also add a pre-computed bool column for the trees in VanDusen Botanical Garden so we don't have to do an expensive filtering step every time the user selects that option in the dropdown.

```{r}
street_trees <- street_trees |>
  mutate(
    # Construct Binomial_Name: capitalize genus and make species lowercase, then remove any trailing " x"
    Binomial_Name = paste0(
      toupper(substr(GENUS_NAME, 1, 1)),
      tolower(substr(GENUS_NAME, 2, nchar(GENUS_NAME))),
      " ",
      tolower(SPECIES_NAME)
    ),
    Binomial_Name = gsub(" x$", "", Binomial_Name),
    Binomial_Name = gsub(" xx$", "", Binomial_Name),
    
    # Convert to title case
    COMMON_NAME = str_to_title(COMMON_NAME),
    NEIGHBOURHOOD_NAME = str_to_title(NEIGHBOURHOOD_NAME),
    CIVIC_ADDRESS = paste0(CIVIC_NUMBER, " ", str_to_title(STD_STREET)),
    
    HEIGHT_RANGE = factor(
      case_when(
        HEIGHT_RANGE == "10-20" ~ "10'-20'",
        HEIGHT_RANGE == "20-30" ~ "20'-30'",
        HEIGHT_RANGE == "30-40" ~ "30'-40'",
        HEIGHT_RANGE == "40-50" ~ "40'-50'",
        HEIGHT_RANGE == "50-60" ~ "50'-60'",
        HEIGHT_RANGE == "60-70" ~ "60'-70'",
        HEIGHT_RANGE == "70-80" ~ "70'-80'",
        HEIGHT_RANGE == "80-90" ~ "80'-90'",
        HEIGHT_RANGE == ">90"   ~ "90'-100'",
        is.na(HEIGHT_RANGE)     ~ NA_character_,
        TRUE                    ~ "Other"
      ),
      levels = c(
        "0'-10'", "10'-20'", "20'-30'", "30'-40'", 
        "40'-50'", "50'-60'", "60'-70'", "70'-80'", 
        "80'-90'", "90'-100'", ">100'"
      ),
      ordered = TRUE
    ),
    
    vandusen_botanical_gardens = LONGITUDE >= -123.138048600311 &
                             LONGITUDE <= -123.12791305292048 &
                             LATITUDE  >= 49.23785042226124 &
                             LATITUDE  <= 49.241214933282045
  )
```

What does it look like?

```{r}
head(street_trees, 10)
tail(street_trees, 10)
```
Let's get rid of additional columns we don't need

```{r}
street_trees <- street_trees |> select(-CIVIC_NUMBER, -STD_STREET, -GENUS_NAME, -SPECIES_NAME)
head(street_trees)
```

Let's look at some VanDusen trees:

```{r}
street_trees |> filter(vandusen_botanical_gardens) |> head()
```

Let's save it out to a binary file so it's fast to read in when the app runs. We don't want to compress it, because this dataset isn't very big and it'll slow down read operations a bit.

```{r}
# Save to fst
write_fst(street_trees, "../data/processed/street-trees.fst", compress = 0)
```

Let's test reading it back in:

```{r}
read_fst("../data/processed/street-trees.fst") |> head()
```
